Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
spell check

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

Although trivial, the previous example captures the fundamental steps of our
construction. For reference, these steps are:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the left
  hand side of each machine state; 
\item Expand the output to expose enough structure to distinguish the right
  hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. This is always a
  mechanical process completely independant of the logic of the machine.
\end{enumerate}

@@

\paragraph*{Stuck States.}
All abstract machines, not just those written in {{langRevT}}, have to
deal with stuck states. There are several standard ways of dealing
with them:
\begin{itemize}
\item \emph{Use better types.}  One could augment {{langRevT}} with a
  richer type system that distinguishes non-zero numbers from those
  that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular intepreter in
  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
  states.

@@

We show the generality of our construction by applying it to two non-trivial
examples: a tree traversal and a meta-circulator interpreter for
{{langRevT}}.

@@

The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are shown on the left:

@@

The reduction rules on the right traverses a given tree and increments
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

In the final construction we present a meta-circular interpreter for
{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the deriving the corresponding
{{langRevT}} interpreter follows in exactly the same way as
before. Here is a logically reversible small-step abstract machine for
{{langRevT}} and the derived isomorphic interpreter with the
reductions labelled.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the `{{iso(v)}}' application in
  {{rule~1}}. This should be read as `tranform {{v}} according to the
  primitive isomorphism {{iso}}'. Implementing this is straightformard
  and is hence skipped.

\item Stuck states in this machine correspond to runtime values not
  matching their expected types. They can be handled as decribed in
  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
  divergence, or adding extra halting states. We have abtracted from
  this choice and marked the relevant cases with combinators labelled
  `{{cast}}'.

@@

We have developed a technique for the systematic development of
{{langRevT}} programs from logically reversible small-step abstract
machines. Since techniques for developing small-step interpreters are
well developed this allows for the direct development of a large class
of {{langRevT}} programs. We have demonstarted the effectiveness of
this approach by deriving meta-circular interpreter for {{langRevT}}
in this style.

>>>>

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of transforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

Although trivial, the previous example captures the fundamental steps of our
construction. For reference, these steps are:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the left
  hand side of each machine state; 
\item Expand the output to expose enough structure to distinguish the right
  hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. This is always a
  mechanical process completely independent of the logic of the machine.
\end{enumerate}

@@

\paragraph*{Stuck States.}
All abstract machines, not just those written in {{langRevT}}, have to
deal with stuck states. There are several standard ways of dealing
with them:
\begin{itemize}
\item \emph{Use better types.}  One could augment {{langRevT}} with a
  richer type system that distinguishes non-zero numbers from those
  that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular interpreter in
  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
  states.

@@

We show the generality of our construction by applying it to two non-trivial
examples: a tree traversal and a meta-circular interpreter for
{{langRevT}}.

@@

The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
sub-tree is currently being explored. The definitions are shown on the left:

@@

The reduction rules on the right traverses a given tree and increments
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored sub-trees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

In the final construction we present a meta-circular interpreter for
{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the deriving the corresponding
{{langRevT}} interpreter follows in exactly the same way as
before. Here is a logically reversible small-step abstract machine for
{{langRevT}} and the derived isomorphic interpreter with the
reductions labeled.

@@

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the `{{iso(v)}}' application in
  {{rule~1}}. This should be read as `transform {{v}} according to the
  primitive isomorphism {{iso}}'. Implementing this is straightforward
  and is hence skipped.

\item Stuck states in this machine correspond to runtime values not
  matching their expected types. They can be handled as described in
  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
  divergence, or adding extra halting states. We have abstracted from
  this choice and marked the relevant cases with combinators labeled
  `{{cast}}'.

@@

We have developed a technique for the systematic development of
{{langRevT}} programs from logically reversible small-step abstract
machines. Since techniques for developing small-step interpreters are
well developed this allows for the direct development of a large class
of {{langRevT}} programs. We have demonstrated the effectiveness of
this approach by deriving meta-circular interpreter for {{langRevT}}
in this style.
